---
title: "Exploratory analysis"
output:
  workflowr::wflow_html:
    toc: true
    code_folding: hide 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

Here we present some exploratory analysis and plots investigating general patterns in the two biodiversity measures estimated at the RainDrop site as part of the Drought-Net experiment, above-ground biomass (`biomass`) and species diversity (`percent_cover`). 

First some housekeeping. We'll start by loading the necessary packages for exploring the data, the data itself and then specifying the colour palette for the treatments.

```{r packages data load colours, fig.width = 5, fig.height = 2}
# packages
library(tidyverse)
library(patchwork)

# load data
load("../../RainDropRobotics/Data/raindrop_biodiversity_2016_2020.RData", 
     verbose = TRUE)

# colours
raindrop_colours <- 
  tibble(treatment = c("Ambient", "Control", "Drought", "Irrigated"),
         num = rep(100,4),
         colour = c("#61D94E", "#BFBFBF", "#EB8344", "#6ECCFF")) 

ggplot(raindrop_colours, aes(x = treatment, y = num, fill = treatment)) +
  geom_col() +
  geom_text(aes(label = colour, x = treatment, y = 50), size = 3) +
  scale_fill_manual(values = raindrop_colours$colour) +
  theme_void()
```

# 1. Above-ground biomass

## Data summary

Here, cuttings of 1m x 0.25m strips (from ~1cm above the ground) from each treatment and block are sorted by functional group and after drying at 70$^\circ$C for over 48 hours their dry mass is weighed in grams. Measurements occur twice per year- one harvest in June in the middle of the growing season and one and in September at the end of the growing season.

Here a summary of the data:

```{r biomass glimpse}
glimpse(biomass)
```

* $**year**      - year of measurement (integer 2016-2020)
* $**harvest**   - point of the growing season when harvest occur (category "Mid" and "End")
* $**block**     - block of experiment (category A-E)
* $**treatment** - experimental treatment for Drought-Net (category "Ambient", "Control", "Drought" and "Irrigated")
* $**group**     - functional group of measurement (category "Bryophytes", "Dead", "Forbs", "Graminoids", "Legumes" and "Woody")
* $**biomass_g** - above-ground dry mass in grams (continuous)

***

## Biomass variable

Let us first have a look at the response variable `biomass_g`. We'll remove any **zero observations** of biomass. If we plot out the frequency histogram of the raw data, we can see there is some strong positive skew in this variable, because of the differences in biomass scale between our functional groups (i.e. grasses very over-represented).

```{r biomass raw hist, fig.width= 5, fig.height= 3}
biomass <- biomass %>% 
  filter(biomass_g > 0)

ggplot(biomass, aes(x = biomass_g)) +
  geom_histogram(bins = 20) +
  labs(x = "Above-ground biomass (g)", y = "Frequency") +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank())
```

So, instead, for statistical analysis and visualisation, we take the natural logarithm $\ln$ of the raw biomass value.

```{r biomass log hist, fig.width= 5, fig.height= 3}
biomass <- biomass %>% 
  mutate(biomass_log = log(biomass_g))

ggplot(biomass, aes(x = biomass_log)) +
  geom_histogram(bins = 20) +
  labs(x = expression(paste(ln," Above-ground biomass")), y = "Frequency") +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank())
```

Using the ln biomass variable, we can explore patterns relating to our drought treatments, functional groups, temporal patterns and any potential block effects.

***

## Total biomass

The first interesting question to explore is how biomass overall, which indicates primary productivity, is affected by the rainfall treatments. Here we need to sum up the biomass across functional groups. We again look at the $\ln$ total biomass too. We present here the density plots (violins) with median (total biomass) and mean ($\ln$ total biomass) points (large holes) as well as underlying raw data (small points).

```{r total biomass}
totbiomass <- biomass %>% 
  group_by(year, harvest, block, treatment) %>% 
  summarise(tot_biomass = sum(biomass_g),
            log_tot_biomass = log(tot_biomass)) %>% 
  ungroup()
```

And now we can look at how this is affected by the rainfall treatments.

```{r total biomass treatment, fig.width= 12, fig.height= 5}
tb_1 <- totbiomass %>% 
  ggplot(aes(x = treatment, y = tot_biomass, 
             fill = treatment)) +
  geom_violin() +
  stat_summary(fun = median, geom = "point", 
               size = 6, shape = 21, fill = "white") +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1)+
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F) +
  labs(x = "Rainfall treatment", y = "Total above-ground biomass (g)") +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank())

tb_2 <- totbiomass %>% 
  ggplot(aes(x = treatment, y = log_tot_biomass, 
             fill = treatment)) +
  geom_violin() +
  stat_summary(fun = mean, geom = "point", 
               size = 6, shape = 21, fill = "white") +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1)+
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F) +
  labs(x = "Rainfall treatment", 
       y = expression(paste(ln, " Total above-ground biomass"))) +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank())

tb_1 + tb_2

```

Across all functional groups, years and blocks, it looks like there is a reduction in total biomass associated with the drought treatment. 

Now we'll look at potential temporal effects i.e. whether there is a change in total biomass over time. We first have to add in temporal information to the year and harvest columns, and then we can plot out the total biomass.

```{r tot biomass temporal, fig.width= 7, fig.height= 5}

# adding temporal information
totbiomass <- totbiomass %>% 
  mutate(month = if_else(harvest == "End", 9, 6),
         date = as.Date(paste0(year,"-",month,"-15")),
         harvest = factor(harvest, levels = c("Mid", "End")))

# plots
totbiomass %>% 
  ggplot(aes(x = date, y = log_tot_biomass, colour = treatment, 
             group = treatment, shape = harvest)) +
  stat_summary(geom = "line", fun = mean, size = 0.9) +
  stat_summary(geom = "point", fun = mean, size = 3) +
  scale_colour_manual(values = raindrop_colours$colour) +
  labs(x = "Year", y = expression(paste("Mean ", ln, " total biomass")),
       colour = "Treatment", shape = "Harvest") +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank())

```

And also splitting out the harvests, as the end of the growing season has lower biomass

```{r tot biomass temporal 2, fig.width= 12, fig.height= 5}
totbiomass %>% 
  ggplot(aes(x = year, y = log_tot_biomass, colour = treatment, 
             group = treatment)) +
  stat_summary(geom = "line", fun = mean, size = 0.9) +
  stat_summary(geom = "point", fun = mean, size = 3) +
  scale_colour_manual(values = raindrop_colours$colour) +
  labs(x = "Year", y = expression(paste("Mean ", ln, " total biomass")),
       colour = "Treatment") +
  facet_wrap(~harvest) +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank(),
        strip.background = element_blank())
```

It looks like there may have been declines in total biomass over the years (perhaps reflecting wider climatic patterns?), but this also seems to be more prominent in the drought treatment.

Finally for total biomass, we want to look if we have any evidence for block effects at the experimental site. Here we will repeat the first treatment violin plot, but splitting by block. Here each violin is a different block (A-E from left to right) with the 50% quantile.

```{r total biomass block, fig.width= 6, fig.height= 5}
totbiomass %>% 
  ggplot(aes(x = treatment, y = log_tot_biomass, 
             fill = treatment, 
             group = interaction(treatment, block))) +
  geom_violin(draw_quantiles = 0.5) +
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F) +
  labs(x = "Rainfall treatment", 
       y = expression(paste(ln, " Total above-ground biomass"))) +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank())
```

It seems that largely blocks are consistent across the RainDrop site.

***

## Group-level biomass

Now we can go a little further and look at whether we observe any differences in biomass between different functional groups, which occur at different proportions. First, lets look at the overall differences in biomass, or relative biomass, between functional groups.

```{r functional group biomass, fig.width= 6, fig.height= 5}
biomass %>% 
  ggplot(aes(x = group, y = biomass_log, 
             fill = group)) +
  geom_violin() +
  stat_summary(fun = mean, geom = "point", 
               size = 5, shape = 21, fill = "white") +
  geom_jitter(width = 0.15, alpha = 0.2, size = 1) +
  scale_fill_brewer(palette = "Set2", guide = F) +
  labs(x = "Functional group", 
       y = expression(paste(ln, " Above-ground biomass"))) +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank())
  
```

We can see that there is a large representation (as expected) of Graminoids, and lower representation of Woody species and Bryophytes.

Now lets look at the rainfall treatments by functional group.

```{r functional group treatment, fig.width= 12, fig.height= 12}
biomass %>% 
  ggplot(aes(x = treatment, y = biomass_log, 
             fill = treatment)) +
  geom_violin() +
  stat_summary(fun = mean, geom = "point", 
               size = 5, shape = 21, fill = "white") +
  geom_jitter(width = 0.15, alpha = 0.2, size = 1) +
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F) +
  labs(x = "Rainfall treatment", 
       y = expression(paste(ln, " Above-ground biomass"))) +
  facet_wrap(~ group, ncol = 2) +
  theme_bw(base_size = 22) +
  theme(panel.grid = element_blank(),
        strip.background = element_blank())
```

It isn't as clear hear as total biomass. However, for the group with the most representation, the Graminoids, this reduction in the drought treatment looks to be maintained.

***

# 2. Species diversity

Shannon-Weiner diversity index per block/treatment/year, we already have relative abundances

```{r}
#H.I <- -sum(pI*log(pI))
```
