---
title: "Mixed effects models"
output:
  workflowr::wflow_html:
    toc: true
    code_folding: hide 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

Here we present the results of Bayesian hierarchical mixed-effects models to explore how the rainfall treatments have influenced biodiversity, namely, the above-ground biomass (both total at the group level) and species diversity (Shannon-Wiener index and species richness) at the RainDrop site. Please refer to the file `code/mixed_effects_models.R` for the full details of the model selection. We'll only be presenting the best predictive models here.

First some housekeeping. We'll start by loading the necessary packages for exploring the data and the modelling, the raw data itself and then specifying the colour palette for the treatments.

```{r packages data load colours, fig.width = 5, fig.height = 2}
# packages
library(tidyverse)
library(brms)
library(tidybayes)
library(patchwork)
library(gghalves)
library(flextable)
library(vegan)

# load data
load("../../RainDropRobotics/Data/raindrop_biodiversity_2016_2020.RData", 
     verbose = TRUE)

# wrangling the data
# Converting to total biomass - Here we are considering TRUE 0 observations - need to chat to Andy about this
totbiomass <- biomass %>%
  group_by(year, harvest, block, treatment) %>%
  summarise(tot_biomass = sum(biomass_g),
            log_tot_biomass = log(tot_biomass + 1)) %>%
  ungroup() %>%
  mutate(year_f = as.factor(year),
         year_s = as.numeric(scale(year)))

## group biomass - Focussing on only graminoids, forbs and legumes
biomass_gr <- biomass %>%
  filter(group %in% c("Forbs", "Graminoids", "Legumes")) %>%
  mutate(biomass_log = log(biomass_g + 1),
         year_f = as.factor(year),
         year_s = as.numeric(scale(year)))


# Percent cover - total biodiversity indices
percent_cover <- percent_cover %>%
  drop_na(percent_cover) %>%
  filter(species_level == "Yes")

diversity <- percent_cover %>%
  # first convert percentages to a proportion
  mutate(proportion = percent_cover/100) %>%
  group_by(year, month, block, treatment) %>%
  # diversity indices for each group
  summarise(richness = n(),
            simpsons = sum(proportion^2),
            shannon = -sum(proportion * log(proportion))) %>%
  ungroup() %>%
  mutate(year_f = as.factor(year),
         year_s = as.numeric(scale(year)),
         shannon = as.numeric(scale(shannon)))

# colours
raindrop_colours <- 
  tibble(treatment = c("Ambient", "Control", "Drought", "Irrigated"),
         num = rep(100,4),
         colour = c("#61D94E", "#BFBFBF", "#EB8344", "#6ECCFF")) 

```

# Modelling specifications

As seen in the [introduction](index.html) and [exploratory analyses](exploratory_analysis.html), the RainDrop study follows a randomised block design with a nested structure and annual sampling. Each treatment plot occurs within one of 5 spatial blocks, and there is temporal replication across years, and so we used hierarchical models with treatment clustered within spatial block and an intercept only random effect of year (as a factor).

The biodiversity response variables used here are:

1. Total biomass - $total\,biomass$ - $\ln$-transformed - Gaussian model
2. Group-level biomass - $biomass$ - $\ln$-transformed - Gaussian model
3. Shannon-Wiener index - $shannon$ - z-transformed - Gaussian model
4. Species richness - $richness$ - Poisson model with a $\log$ link

The full base models and priors are given at the beginning of each subsequent section, with the general priors specified for each. Priors were adjusted for each model to prevent divergent transitions, see `code/mixed_effects_models.R`.

Final models were run over 4 chains, with 4000 iterations of which 2000 were warmup iterations (2000 sampling iterations). We increased the target proposal acceptance probability (i.e. how conservative the acceptance of samples was) of the NUTS sampler by increasing `adapt_delta` (default = 0.8 from `rstan`) >= 0.95 (maximum 0.98) in each model to prevent divergent transitions. Model convergence was assessed with $\hat{R}$.

## Model selection

We wanted to assess how the rainfall treatments influenced each of our biodiversity variable, and also how these patterns may have changed through time. We therefore used leave-one-out cross validation implemented in the `loo` package for a set of candidate models, and assessed differences in estimated out-of-sample predictive performance using model selection. Here, we compared models directly using their expected log pointwise predictive density (elpd). For the full set of models explored for each biodiversity variable, please refer to their specific section.

Generally however, we explored the following additions to the base model:

* The rainfall treatment (`treatment`)
* A Linear year effect (`year_s`)
* A functional group effect (`group`)
* An interaction between treatment and year
* An interaction between functional group and treatment/year
* An interaction between year and harvest point (or collection month)
* A year term with temporal autocorrelation (lag 1) (`ar(year_s, p = 1)`)

## Autocorrelated year effects

Modelling the temporal component to this dataset alone may not effectively account for temporal autocorrelation in biodiversity, and therefore we implemented autocorrelated year effects with a lag of 1. These autocorrelated terms were implemented using the `ar()` function of the `brms` package, in which we specified a lag of 1.

---

# 1. Total biomass

The full base model for total biomass is as follows, where for a biomass observation in block $i$ ($i = 1..5$), treatment $j$ ($j = 1..4$) and year $t$ ($t = 1..5$)

$$
\begin{aligned}
\operatorname{Full model} \\
total\,biomass_{ijt} &\sim \operatorname{Normal}(\mu_{ijt}, \sigma) \\
\mu_{ijt} &= \alpha_{i} + \beta_{harvest[ijt]} + \epsilon_{ij} + \gamma_t  \\
\operatorname{Priors} \\
\beta_{harvest[ijt]} &\sim \operatorname{Normal}(0, 1) \\
\alpha_i &\sim \operatorname{Normal}(\bar{\alpha},\sigma_{\alpha}) \\
\epsilon_{ij} &\sim \operatorname{Normal}(0,\sigma_{\epsilon})\\
\gamma_t &\sim \operatorname{Normal}(0,\sigma_{\gamma})\\
\bar{\alpha} &\sim\operatorname{Normal}(3.5, 0.5)\\
\sigma_{\alpha} &\sim \operatorname{Exponential}(3) \\
\sigma_{\epsilon} &\sim \operatorname{Exponential}(3) \\
\sigma_{\gamma} &\sim \operatorname{Exponential}(3) \\
\sigma &\sim \operatorname{Student T}(3,0,2.5) \\
\end{aligned}
$$

Where $\epsilon_{ij}$ and $\gamma_t$ are intercept-only random effects for the cluster of treatment within block and year, respectively.

## Model selection results

Now we load the results from the model selection and the model with the highest predictive performance and have a look at the results of the model selection. Model terms are additional to those in the base model.

```{r tot biomass results}
load("data/totbiomass_models.RData", verbose = TRUE)

## model selection table
round(x = mod_comp_totbiomass, 2) %>% 
  rownames_to_column("R object") %>% 
  mutate(`Model terms` = c("treatment", "treatment + year_s",
                           "treatment + year_s + harvest:year_s",
                           "treatment + year_s + treatment:year_s",
                           "base model",
                           "treatment + ar(year_s, p = 1)")) %>% 
  dplyr::select(`R object`, `Model terms`, `LOO elpd` = elpd_loo, 
                `LOO elpd error` = se_elpd_loo, `elpd difference` = elpd_diff,
                `elpd error difference` = se_diff, `LOO information criterion` = looic) %>% 
  flextable(cwidth = 1.5)

```

We can see that the model with the highest predictive performance for total biomass is that with just the treatment effect. There is evidence for a linear year effect and interactions between year and treatment/harvest relative to the base model, but these are not more predictive than just the treatment effect. We have poor support for temporal autocorrelation in the total biomass.

## Posterior plots and predictions

First lets have a look at the terms not added after model selection. Here we present distributions from 2000 draws from the posterior of the best model.

```{r posterior tot biomass, fig.width= 8, fig.height = 9}
totbiomass_treatment %>%
  gather_draws(`b_harvestMid|sd.*|sigma`, regex = TRUE) %>% #tidybayes
  ungroup() %>% 
  ggplot(aes(y = .variable, x = .value)) + 
  stat_halfeye(show.legend = FALSE, fill = "lightsteelblue") +
  scale_y_discrete(labels = c(expression(paste("Early harvest effect ", beta[harvest])),
                              expression(paste("Treatment within block variance ", sigma[epsilon])),
                              expression(paste("Block variance ", sigma[alpha])),
                              expression(paste("Year variance ", sigma[gamma])),
                              expression(paste("Population-level variance ", sigma)))) +
  labs(x = "Posterior estimate", y = NULL) +
  theme_bw(base_size = 20)
```

We can see that first of all we have a strong positive effect of the Harvest on total biomass, where biomass is higher by over 50% compared to the later growing season. Then, for the variance terms, we can see that relative to the population-level i.e. total variance, block-level and treatment within block variances were low. In contrast, there was high inter-annual variance.

Now, for the treatment effect. Here we present the mean distribution of posterior predictions for each of the treatment groups and the raw data.

```{r post pred tot biomass, fig.width= 9, fig.height = 6}
# Harvest labels
harvest.labels <- c("End of season harvest", "Mid season Harvest")
names(harvest.labels) <- c("End", "Mid")

as.data.frame(brms::posterior_predict(totbiomass_treatment)) %>% 
  mutate(sim = 1:8000) %>% 
  pivot_longer(-sim) %>% 
  bind_cols(., slice(totbiomass, rep(1:200, 8000))) %>% 
  mutate(harvest = factor(harvest, levels = c("Mid", "End"))) %>% 
  ggplot(aes(x = treatment, y = value, fill = treatment, group = harvest)) +
  stat_halfeye(width = 0.4, alpha = 0.7) +
  geom_half_point(data = totbiomass, 
                  aes(y = log_tot_biomass, x = treatment, 
                      colour = treatment, group = NULL),
                  side = "l", ## draw jitter on the left
                  range_scale = 0.4, ## control range of jitter
                  alpha = 0.7, size = 2) +
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F, aesthetics = c("fill", "colour")) +
  labs(x = "Rainfall treatment", y = expression(paste(ln, " Total biomass"))) +
  facet_wrap(~ harvest, ncol = 2,
             labeller = labeller(harvest = harvest.labels),) +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank(),
        strip.background = element_blank())
  
```

From this it is clear that there is a **reduction in total biomass associated with the drought treatment**. There doesn't seem to be any clear differences between ambient, control or irrigated plots however. These differences need to be tested further as well patterns in the variance.

## Hypothesis testing

We wanted to test a few key hypotheses

### 1. What are the ICCs for both the clustered and year random effects

How does the variance in our random effect terms compare to the total variance? An example of this calculation for the block within treatment variance is below

$$ICC_{\sigma_{\epsilon}} = \frac {\sigma^{2}_{\epsilon}}{\sigma^{2}_{\epsilon} + \sigma^{2}_{\alpha} + \sigma^{2}_{\gamma} + \sigma}$$
We test the hypothesis of whether this ICC value is equal to zero. First for the treatment within block variance $\sigma_{\epsilon}$

```{r sigma epsilon ICC}
h <- paste("sd_block:treatment__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(totbiomass_treatment, h, class = NULL)
```

Now the block variance

```{r sigma alpha ICC}
h <- paste("sd_block__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(totbiomass_treatment, h, class = NULL)
```

Finally our inter-annual variance

```{r sigma gamma ICC}
h <- paste("sd_year_f__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(totbiomass_treatment, h, class = NULL)
```

You can see we have much higher intraclass correlation with the interannual variance, a repeatable year-level biomass, but no clear ICC in our block/treatment effects.

### 2. Is the difference in biomass between the control treatments is > 0?

Now, we want to test whether the different control plots have a substantially different biomass. Here we are simply testing whether their difference is equal to 0.

```{r hyp controls}
hypothesis(totbiomass_treatment, "Intercept - treatmentControl = 0")
```

### 3. Evidence for an effect of irrigation on biomass relative to ambient?
 
And finally for any evidence of an irrigation effect

```{r hyp irrigation}
hypothesis(totbiomass_treatment, "Intercept - treatmentIrrigated = 0")
```

---

# 2. Group-level biomass

Next we explored the biomass differences between treatments for the three key functional groups: Graminoids, Legumes and Forbs. The full base model here is the same as for total biomass:

$$
\begin{aligned}
\operatorname{Full model} \\
biomass_{ijt} &\sim \operatorname{Normal}(\mu_{ijt}, \sigma) \\
\mu_{ijt} &= \alpha_{i} + \beta_{harvest[ijt]} + \epsilon_{ij} + \gamma_t  \\
\operatorname{Priors} \\
\beta_{harvest[ijt]} &\sim \operatorname{Normal}(0, 0.5) \\
\alpha_i &\sim \operatorname{Normal}(\bar{\alpha},\sigma_{\alpha}) \\
\epsilon_{ij} &\sim \operatorname{Normal}(0,\sigma_{\epsilon})\\
\gamma_t &\sim \operatorname{Normal}(0,\sigma_{\gamma})\\
\bar{\alpha} &\sim\operatorname{Normal}(1, 1)\\
\sigma_{\alpha} &\sim \operatorname{Exponential}(3) \\
\sigma_{\epsilon} &\sim \operatorname{Exponential}(3) \\
\sigma_{\gamma} &\sim \operatorname{Exponential}(3) \\
\sigma &\sim \operatorname{Student T}(3,0,2.5) \\
\end{aligned}
$$

## Model selection results

Now we load the results from the model selection and the model with the highest predictive performance and have a look at the results of the model selection. Model terms are additional to those in the base model. For this group-level biomass, we also included terms relating to the functional group (`group`) of the biomass measure, and appropriate interactions.

```{r biomass gr mod comp}
load("data/biomass_models.RData", verbose = TRUE)

## model selection table
round(x = mod_comp_biomass, 2) %>% 
  rownames_to_column("R object") %>% 
  mutate(`Model terms` = c("treatment + group + treatment:group",
                           "treatment + group + year_s + treatment:group + treatment:year_s + group:year_s + treatment:group:year_s",
                           "treatment + group",
                           "group",
                           "treatment",
                           "base model")) %>% 
  dplyr::select(`R object`, `Model terms`, `LOO elpd` = elpd_loo, 
                `LOO elpd error` = se_elpd_loo, `elpd difference` = elpd_diff,
                `elpd error difference` = se_diff, `LOO information criterion` = looic) %>% 
  flextable(cwidth = 1.5)
```

Here the model with the highest predictive performance is the one with a two way interaction between rainfall treatment and functional group. This model has more clear statistical support, with an elpd difference of greater than 3 compared to the second most competitive model.

## Posterior plots and predictions

First lets have a look at the terms not added after model selection. Here we present distributions from 2000 draws from the posterior of the best model.

```{r posterior biomass gr, fig.width= 8, fig.height = 9}
biomass_treatment_group_int %>%
  gather_draws(`b_harvestMid|sd.*|sigma`, regex = TRUE) %>% #tidybayes
  ungroup() %>% 
  ggplot(aes(y = .variable, x = .value)) + 
  stat_halfeye(show.legend = FALSE, fill = "lightsteelblue2") +
  scale_y_discrete(labels = c(expression(paste("Early harvest effect ", beta[harvest])),
                              expression(paste("Treatment within block variance ", sigma[epsilon])),
                              expression(paste("Block variance ", sigma[alpha])),
                              expression(paste("Year variance ", sigma[gamma])),
                              expression(paste("Population-level variance ", sigma)))) +
  labs(x = "Posterior estimate", y = NULL) +
  theme_bw(base_size = 20)
```

Again we have a strong positive effect of the Harvest on biomass, regardless of functional group, with a similar coefficient. Then, for the variance terms, we see that block/treatment variance terms remain low, but with treatment within block variance slightly higher. Year variance remained high.

Now, for the treatment and group effects. Here we present the mean distribution of posterior predictions for each of the treatment groups and the raw data.

```{r biomass group predictions, fig.width= 13, fig.height = 6}
as.data.frame(brms::posterior_predict(biomass_treatment_group_int, nsamples = 2000)) %>% # reducing to make data more manageable 
  mutate(sim = 1:2000) %>% 
  pivot_longer(-sim) %>% 
  bind_cols(., slice(biomass_gr, rep(1:600, 2000))) %>% 
  filter(harvest == "Mid") %>% 
  ggplot(aes(x = treatment, y = value, fill = treatment, group = group)) +
  stat_halfeye(width = 0.4, alpha = 0.7) +
  geom_half_point(data = biomass_gr, 
                  aes(y = biomass_log, x = treatment, 
                      colour = treatment, group = NULL),
                  side = "l", ## draw jitter on the left
                  range_scale = 0.4, ## control range of jitter
                  alpha = 0.7, size = 2) +
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F, aesthetics = c("fill", "colour")) +
  labs(x = "Rainfall treatment", y = expression(paste(ln, " Total biomass"))) +
  facet_wrap(~ group, ncol = 3) +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank(),
        strip.background = element_blank())
```

Here we have some very interesting differences in biomass between the functional groups in response to the rainfall treatments. First of all, it is clear that Graminoids are more prevalent than other groups, with $\beta_{group, graminoid} = 1.39 [1.15,1.64]$. However, we see the negative impact of the drought treatment on biomass is only present in the Graminoids and Legumes, but not the Forbs, which showed little response to the treatments. There is also a slight indication that the Irrigated treatment had reduced Legume biomass $$\beta_{legume:irrigated} = -0.18 [-0.52,0.16]$$, but this had weaker statistical support.

## Hypothesis testing

### 1. What are the ICCs for both the clustered and year random effects

Treatment within block variance - slightly higher but still little support

```{r sigma epsilon ICC gr}
h <- paste("sd_block:treatment__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(biomass_treatment_group_int, h, class = NULL)
```

Block variance

```{r sigma alpha ICC gr}
h <- paste("sd_block__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(biomass_treatment_group_int, h, class = NULL)
```

Inter-annual variance - stronger annual variance effect

```{r sigma gamma ICC gr}
h <- paste("sd_year_f__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(biomass_treatment_group_int, h, class = NULL)
```

### 2. Is the difference in biomass between the control treatments is > 0?

```{r hyp controls gr}
hypothesis(biomass_treatment_group_int, "Intercept - treatmentControl = 0")
```

---

# 3. Shannon-Weiner diversity index

Now we wanted to get handle on species diversity, starting first with the Shannon-Weiner index. Here we have month instead of harvest (June or September) as it wasn't a harvest.

$$
\begin{aligned}
\operatorname{Full model} \\
shannon_{ijt} &\sim \operatorname{Normal}(\mu_{ijt}, \sigma) \\
\mu_{ijt} &= \alpha_{i} + \beta_{month[ijt]} + \epsilon_{ij} + \gamma_t  \\
\operatorname{Priors} \\
\beta_{month[ijt]} &\sim \operatorname{Normal}(0, 1) \\
\alpha_i &\sim \operatorname{Normal}(\bar{\alpha},\sigma_{\alpha}) \\
\epsilon_{ij} &\sim \operatorname{Normal}(0,\sigma_{\epsilon})\\
\gamma_t &\sim \operatorname{Normal}(0,\sigma_{\gamma})\\
\bar{\alpha} &\sim\operatorname{Normal}(3.5, 0.5)\\
\sigma_{\alpha} &\sim \operatorname{Exponential}(3) \\
\sigma_{\epsilon} &\sim \operatorname{Exponential}(3) \\
\sigma_{\gamma} &\sim \operatorname{Exponential}(3) \\
\sigma &\sim \operatorname{Student T}(3,0,2.5) \\
\end{aligned}
$$

## Model selection results

The model selection took a similar form to previous model selections.

```{r shannon mod comp}
load("data/shannon_models.RData", verbose = TRUE)

## model selection table
round(x = mod_comp_shannon, 2) %>% 
  rownames_to_column("R object") %>% 
  mutate(`Model terms` = c("treatment + year_s + treatment:year_s",
                           "base model",
                           "treatment + year_s",
                           "treatment",
                           "treatment + ar(year_s, p = 1)")) %>% 
  dplyr::select(`R object`, `Model terms`, `LOO elpd` = elpd_loo, 
                `LOO elpd error` = se_elpd_loo, `elpd difference` = elpd_diff,
                `elpd error difference` = se_diff, `LOO information criterion` = looic) %>% 
  flextable(cwidth = 1.5)
```

The best model here was the one with an interaction between a linear year effect and the treatment group, the only candidate model better than the base model. This model had strong statistical support however, with an elpd difference greater than 5. Again we found little evidence for an auto-correlated year term.

## Posterior plots and predictions

First we present the terms in the base model again. 

```{r posterior shannon, fig.width= 8, fig.height = 9}
shannon_treatment_year_int %>%
  gather_draws(`b_monthSept|sd.*|sigma`, regex = TRUE) %>% #tidybayes
  ungroup() %>% 
  ggplot(aes(y = .variable, x = .value)) + 
  stat_halfeye(show.legend = FALSE, fill = "lightsteelblue4") +
  scale_y_discrete(labels = c(expression(paste("September effect ", beta[harvest])),
                              expression(paste("Treatment within block variance ", sigma[epsilon])),
                              expression(paste("Block variance ", sigma[alpha])),
                              expression(paste("Year variance ", sigma[gamma])),
                              expression(paste("Population-level variance ", sigma)))) +
  labs(x = "Posterior estimate", y = NULL) +
  theme_bw(base_size = 20)
```

In terms of species diversity, we see some large differences compared to biomass. First, the treatment within block variance was much larger in this measure, suggesting that communities on the site are more localised. There also seems to be less of a variance component for year. Further, There isn't a clear change in species diversity depending on the observation month.


```{r shannon predictions, fig.width= 8, fig.height = 7}
preddat_shannon <- expand_grid(year_s = seq(-1.15,1.6, length.out = 40),
                               treatment = unique(diversity$treatment),
                               month = "June",
                               block = unique(diversity$block),
                               year_f = unique(diversity$year_f),
                               value = NA) 

shannon_pred <-  brms::posterior_predict(shannon_treatment_year_int,
                                         newdata = preddat_shannon) 

# summary data for each year value
predatsum <- preddat_shannon %>% 
  group_by(year_s, treatment) %>% 
  summarise(value = NA, upr = NA, lwr = NA) %>% 
  ungroup() %>% as.data.frame()
  
for(i in 1:nrow(predatsum)){
  cpos = which(preddat_shannon$year_s == predatsum[i,]$year_s & 
                 preddat_shannon$treatment == predatsum[i,]$treatment)
  
  post_mean = mean(shannon_pred[,cpos])
  cQuant = quantile(shannon_pred[,cpos], c(0.1, 0.90)) # 80% levels
  
  predatsum[i,]$value <- post_mean
  predatsum[i,]$upr <- cQuant[2]
  predatsum[i,]$lwr <- cQuant[1]
  
}

ggplot(predatsum, aes(x = year_s, y = value, group = treatment, 
                      fill = treatment, colour = treatment)) +
  geom_jitter(data = diversity, aes(y = shannon), width = 0.06, size = 2, alpha = 0.85) +
  geom_smooth(stat = "identity", aes(ymax = upr, ymin = lwr), alpha = 0.1) +
  scale_fill_manual(values = raindrop_colours$colour,
                    aesthetics = c("fill", "colour"), name = "Rainfall\ntreatment") +
  labs(x = "Observation year (scaled)", y = "Shannon-Weiner index") +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank())

```

Here, we have a very interesting observation that suggests that as the study has progressed, the Shannon-Weiner diversity in the Irrigated treatments has decreased, but not in other treatment groups. Could this suggest a movement towards an increased monoculture with excessive rainfall?

## Hypothesis testing

Here we just want to test the ICC values.

### 1. What are the ICCs for both the clustered and year random effects

Treatment within block variance - Here this value is the greatest, highlighting the results from before.

```{r sigma epsilon ICC sh}
h <- paste("sd_block:treatment__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(shannon_treatment_year_int, h, class = NULL)
```

Block variance

```{r sigma alpha ICC sh}
h <- paste("sd_block__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(shannon_treatment_year_int, h, class = NULL)
```

Inter-annual variance 

```{r sigma gamma ICC sh}
h <- paste("sd_year_f__Intercept^2/ ", 
           "(sd_block:treatment__Intercept^2 + ",
           "sd_block__Intercept^2 + ",
           "sd_year_f__Intercept^2 + sigma) = 0")

hypothesis(shannon_treatment_year_int, h, class = NULL)
```

---

# 4. Species richness

As well as the Shannon-Weiner diversity index, we also wanted to explore patterns in species richness. This variable is count data (no. species) and thus we used a Poisson model with a $\log$ link

$$
\begin{aligned}
\operatorname{Full model} \\
richness_{ijt} &\sim \operatorname{Poisson}(\lambda_{ijt}) \\
\log(\lambda_{ijt}) &= \alpha_{i} + \beta_{month[ijt]} + \epsilon_{ij} + \gamma_t  \\
\operatorname{Priors} \\
\beta_{month[ijt]} &\sim \operatorname{Normal}(0, 1) \\
\alpha_i &\sim \operatorname{Normal}(\bar{\alpha},\sigma_{\alpha}) \\
\epsilon_{ij} &\sim \operatorname{Normal}(0,\sigma_{\epsilon})\\
\gamma_t &\sim \operatorname{Normal}(0,\sigma_{\gamma})\\
\bar{\alpha} &\sim\operatorname{Normal}(1, 1)\\
\sigma_{\alpha} &\sim \operatorname{Exponential}(3) \\
\sigma_{\epsilon} &\sim \operatorname{Exponential}(3) \\
\sigma_{\gamma} &\sim \operatorname{Exponential}(3) \\
\end{aligned}
$$

## Model selection results

The model selection took a similar form to previous model selections.

```{r richness mod comp}
load("data/richness_models.RData", verbose = TRUE)

## model selection table
round(x = mod_comp_richness, 2) %>% 
  rownames_to_column("R object") %>% 
  mutate(`Model terms` = c("treatment + year_s",
                           "treatment",
                           "base model",
                           "treatment + ar(year_s, p = 1)",
                           "treatment + year_s + treatment:year_s")) %>% 
  dplyr::select(`R object`, `Model terms`, `LOO elpd` = elpd_loo, 
                `LOO elpd error` = se_elpd_loo, `elpd difference` = elpd_diff,
                `elpd error difference` = se_diff, `LOO information criterion` = looic) %>% 
  flextable(cwidth = 1.5)
```

Here we see the strongest statistical support for the model with year and treatment as univariate effects i.e. excluding interactions. However, this result is less convincing, without a substantial change compared to the base model (difference = 1.02). We conclude that there isn't a strong impact of treatment on richness, or strong changes through time. However, it is still worth exploring these effects.

## Posterior plots and predictions

First again for the base model terms

```{r posterior richness, fig.width= 8, fig.height = 8}
richness_year_linear %>%
  gather_draws(`b_monthSept|sd.*|sigma`, regex = TRUE) %>% #tidybayes
  ungroup() %>% 
  ggplot(aes(y = .variable, x = .value)) + 
  stat_halfeye(show.legend = FALSE, fill = "paleturquoise4") +
  scale_y_discrete(labels = c(expression(paste("September effect ", beta[harvest])),
                              expression(paste("Treatment within block variance ", sigma[epsilon])),
                              expression(paste("Block variance ", sigma[alpha])),
                              expression(paste("Year variance ", sigma[gamma])))) +
  labs(x = "Posterior estimate", y = NULL) +
  theme_bw(base_size = 20)
```

Here we don't see strong variance effects for any of the random terms. There is a slight negative effect of the late season on richness, but this isn't that strong. 

Now for the treatment and year effects.

```{r richness predictions, fig.width= 10, fig.height = 5}
# Treatment
p1 <- as.data.frame(brms::posterior_predict(richness_year_linear)) %>% 
  mutate(sim = 1:8000) %>% 
  pivot_longer(-sim) %>% 
  bind_cols(., slice(diversity, rep(1:120, 8000))) %>%
  ggplot(aes(x = treatment, y = value, fill = treatment)) +
  stat_halfeye(width = 0.4, alpha = 0.7) +
  geom_half_point(data = diversity, 
                  aes(y = richness, x = treatment, 
                      colour = treatment, group = NULL),
                  side = "l", range_scale = 0.4, 
                  alpha = 0.7, size = 2) +
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F, aesthetics = c("fill", "colour")) +
  labs(x = "Rainfall treatment", y = "Species Richness") +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank(),
        strip.background = element_blank())

# Year effect
preddat_richness <- expand_grid(year_s = seq(-1.15,1.6, length.out = 40),
                               treatment = unique(diversity$treatment),
                               month = "June",
                               block = unique(diversity$block),
                               year_f = unique(diversity$year_f),
                               value = NA) 

richness_pred <-  brms::posterior_predict(richness_year_linear,
                                         newdata = preddat_richness) 

# summary data for each year value
predatsum <- preddat_richness %>% 
  group_by(year_s) %>% 
  summarise(value = NA, upr = NA, lwr = NA) %>% 
  ungroup() %>% as.data.frame()
  
for(i in 1:nrow(predatsum)){
  cpos = which(preddat_richness$year_s == predatsum[i,]$year_s)
  
  post_mean = mean(richness_pred[,cpos])
  cQuant = quantile(richness_pred[,cpos], c(0.1, 0.90)) # 80% levels
  
  predatsum[i,]$value <- post_mean
  predatsum[i,]$upr <- cQuant[2]
  predatsum[i,]$lwr <- cQuant[1]
  
}

p2 <- ggplot(predatsum, aes(x = year_s, y = value)) +
  geom_jitter(data = diversity, aes(y = richness), width = 0.06, size = 2, alpha = 0.85) +
  geom_smooth(stat = "identity", aes(ymax = upr, ymin = lwr), alpha = 0.1,
              colour = "lightsteelblue4", fill = "lightsteelblue4") +
  labs(x = "Observation year (scaled)", y = "Species richness") +
  theme_bw(base_size = 20) +
  theme(panel.grid = element_blank())

p1 + p2

```

There does look like there was an increase in richness over the study period, although $\beta_{year} = 0.1 [0.04,0.17]$ and this was not a strong effect (particularly in light of predictive performance in the model selection). However, it does look like on average ~10 more species were observed on the site from 2016-2020. There doesn't seem to be any strong effect of the treatment groups.

