---
title: "Mixed effects models"
output:
  workflowr::wflow_html:
    toc: true
    code_folding: hide 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

Here we present the results of Bayesian hierarchical mixed-effects models to explore how the rainfall treatments have influenced biodiversity, namely, the above-ground biomass (both total at the group level) and species diversity (Shannon-Wiener index and species richness) at the RainDrop site. Please refer to the file `code/mixed_effects_models.R` for the full details of the model selection. We'll only be presenting the best predictive models here.

First some housekeeping. We'll start by loading the necessary packages for exploring the data and the modelling, the raw data itself and then specifying the colour palette for the treatments.

```{r packages data load colours, fig.width = 5, fig.height = 2}
# packages
library(tidyverse)
library(brms)
library(tidybayes)
library(patchwork)
library(gghalves)
library(flextable)
library(vegan)

# load data
load("../../RainDropRobotics/Data/raindrop_biodiversity_2016_2020.RData", 
     verbose = TRUE)

# wrangling the data
# Converting to total biomass - Here we are considering TRUE 0 observations - need to chat to Andy about this
totbiomass <- biomass %>%
  group_by(year, harvest, block, treatment) %>%
  summarise(tot_biomass = sum(biomass_g),
            log_tot_biomass = log(tot_biomass + 1)) %>%
  ungroup() %>%
  mutate(year_f = as.factor(year),
         year_s = as.numeric(scale(year)))

## group biomass - Focussing on only graminoids, forbs and legumes
biomass_gr <- biomass %>%
  filter(group %in% c("Forbs", "Graminoids", "Legumes")) %>%
  mutate(biomass_log = log(biomass_g + 1),
         year_f = as.factor(year),
         year_s = as.numeric(scale(year)))


# Percent cover - total biodiversity indices
percent_cover <- percent_cover %>%
  drop_na(percent_cover) %>%
  filter(species_level == "Yes")

diversity <- percent_cover %>%
  # first convert percentages to a proportion
  mutate(proportion = percent_cover/100) %>%
  group_by(year, month, block, treatment) %>%
  # diversity indices for each group
  summarise(richness = n(),
            simpsons = sum(proportion^2),
            shannon = -sum(proportion * log(proportion))) %>%
  ungroup() %>%
  mutate(year_f = as.factor(year),
         year_s = as.numeric(scale(year)),
         shannon = as.numeric(scale(shannon)))

# colours
raindrop_colours <- 
  tibble(treatment = c("Ambient", "Control", "Drought", "Irrigated"),
         num = rep(100,4),
         colour = c("#61D94E", "#BFBFBF", "#EB8344", "#6ECCFF")) 

```

# Modelling specifications

As seen in the [introduction](index.html) and [exploratory analyses](exploratory_analysis.html), the RainDrop study follows a randomised block design with a nested structure and annual sampling. Each treatment plot occurs within one of 5 spatial blocks, and there is temporal replication across years, and so we used hierarchical models with treatment clustered within spatial block and an intercept only random effect of year (as a factor).

The biodiversity response variables used here are:

1. Total biomass - $total\,biomass$ - $\ln$-transformed - Gaussian model
2. Group-level biomass - $biomass$ - $\ln$-transformed - Gaussian model
3. Shannon-Wiener index - $shannon$ - z-transformed - Gaussian model
4. Species richness - $richness$ - Poisson model with a $\log$ link

The full base models and priors are given at the beginning of each subsequent section, with the general priors specified for each. Priors were adjusted for each model to prevent divergent transitions, see `code/mixed_effects_models.R`.

Final models were run over 4 chains, with 4000 iterations of which 2000 were warmup iterations (2000 sampling iterations). We increased the target proposal acceptance probability (i.e. how conservative the acceptance of samples was) of the NUTS sampler by increasing `adapt_delta` (default = 0.8 from `rstan`) >= 0.95 (maximum 0.98) in each model to prevent divergent transitions. Model convergence was assessed with $\hat{R}$.

## Model selection

We wanted to assess how the rainfall treatments influenced each of our biodiversity variable, and also how these patterns may have changed through time. We therefore used leave-one-out cross validation implemented in the `loo` package for a set of candidate models, and assessed differences in estimated out-of-sample predictive performance using model selection. Here, we compared models directly using their expected log pointwise predictive density (elpd). For the full set of models explored for each biodiversity variable, please refer to their specific section.

Generally however, we explored the following additions to the base model:

* The rainfall treatment (`treatment`)
* A Linear year effect (`year_s`)
* A functional group effect (`group`)
* An interaction between treatment and year
* An interaction between functional group and treatment/year
* An interaction between year and harvest point
* A year term with temporal autocorrelation (lag 1) (`ar(year_s, p = 1)`)

## Autocorrelated year effects

Modelling the temporal component to this dataset alone may not effectively account for temporal autocorrelation in biodiversity, and therefore we implemented autocorrelated year effects with a lag of 1. These autocorrelated terms were implemented using the `ar()` function of the `brms` package, in which we specified a lag of 1.

# 1. Total biomass

The full base model for total biomass is as follows, where for a biomass observation in block $i$ ($i = 1..5$), treatment $j$ ($j = 1..4$) and year $t$ ($t = 1..5$)

$$
\begin{aligned}
\operatorname{Full model} \\
total\,biomass_{ijt} &\sim \operatorname{Normal}(\mu_{ijt}, \sigma) \\
\mu_{ijt} &= \alpha_{i} + \beta_{harvest[ijt]} + \epsilon_{ij} + \gamma_t  \\
\operatorname{Priors} \\
\beta_{harvest[ijt]} &\sim \operatorname{Normal}(0, 1) \\
\alpha_i &\sim \operatorname{Normal}(\bar{\alpha},\sigma_{\alpha}) \\
\epsilon_{ij} &\sim \operatorname{Normal}(0,\sigma_{\epsilon})\\
\gamma_t &\sim \operatorname{Normal}(0,\sigma_{\gamma})\\
\bar{\alpha} &\sim\operatorname{Normal}(3.5, 0.5)\\
\sigma_{\alpha} &\sim \operatorname{Exponential}(3) \\
\sigma_{\epsilon} &\sim \operatorname{Exponential}(3) \\
\sigma_{\gamma} &\sim \operatorname{Exponential}(3) \\
\end{aligned}
$$

Where $\epsilon_{ij}$ and $\gamma_t$ are intercept-only random effects for the cluster of treatment within block and year, respectively.

## Model selection results

Now we load the results from the model selection and the model with the highest predictive performance and have a look at the results of the model selection. Model terms are additional to those in the base model.

```{r tot biomass results}
load("data/totbiomass_models.RData", verbose = TRUE)

## model selection table
round(x = mod_comp_totbiomass, 2) %>% 
  rownames_to_column("R object") %>% 
  mutate(`Model terms` = c("treatment", "treatment + year_s",
                           "treatment + year_s + harvest:year_s",
                           "treatment + year_s + treatment:year_s",
                           "base model",
                           "treatment + ar(year_s, p = 1)")) %>% 
  dplyr::select(`R object`, `Model terms`, `LOO elpd` = elpd_loo, 
                `LOO elpd error` = se_elpd_loo, `elpd difference` = elpd_diff,
                `elpd error difference` = se_diff, `LOO information criterion` = looic) %>% 
  flextable(cwidth = 1.5)

```

We can see that the model with the highest predictive performance for total biomass is that with just the treatment effect. There is evidence for a linear year effect and interactions between year and treatment/harvest relative to the base model, but these are not more predictive than just the treatment effect. We have poor support for temporal autocorrelation in the total biomass.

## Posterior plots and predictions

First lets have a look at the terms within the base model. Here we present distributions from 2000 draws from the posterior of the best model.

```{r posterior tot biomass}
totbiomass_treatment %>%
  gather_draws(`b_harvestMid|sd.*|sigma`, regex = TRUE) %>% #tidybayes
  ungroup() %>% 
  ggplot(aes(y = .variable, x = .value)) + 
  stat_halfeye(show.legend = FALSE, fill = "lightsteelblue") +
  scale_y_discrete(labels = c(expression(paste("Early harvest effect ", beta[harvest])),
                              expression(paste("Block within treatment variance ", sigma[epsilon])),
                              expression(paste("Block variance ", sigma[alpha])),
                              expression(paste("Year variance ", sigma[gamma])),
                              expression(paste("Population-level variance ", sigma)))) +
  labs(x = "Posterior estimate", y = NULL) +
  theme_bw(base_size = 15)
```

We can see that first of all we have a strong positive effect of the Harvest on total biomass, where biomass is higher by over 50% compared to the later growing season. Then, for the variance terms, we can see that relative to the population-level i.e. total variance, block-level and treatment within block variances were low. In contrast, there was high inter-annual variance.

Now, for the treatment effect. Here we present the mean distribution of posterior predictions for each of the treatment groups and the raw data.

```{r post pred tot biomass}
as.data.frame(brms::posterior_predict(totbiomass_treatment)) %>% 
  mutate(sim = 1:8000) %>% 
  pivot_longer(-sim) %>% 
  bind_cols(., slice(totbiomass, rep(1:200, 8000))) %>% 
  group_by(treatment, sim) %>% 
  summarise(mn_pred = mean(value)) %>% 
  ggplot(aes(x = treatment, y = mn_pred, fill = treatment)) +
  stat_halfeye(width = 0.4) +
  geom_half_point(data = totbiomass, 
                  aes(y = log_tot_biomass, colour = treatment),
                  side = "l", ## draw jitter on the left
                  range_scale = 0.4, ## control range of jitter
                  alpha = 0.7) +
  scale_fill_manual(values = raindrop_colours$colour,
                    guide = F, aesthetics = c("fill", "colour")) +
  labs(x = "Rainfall treatment", y = expression(paste(ln, " Total biomass"))) +
  theme_bw(base_size = 15) +
  theme(panel.grid = element_blank())
  
```




# 2. Group-level biomass


